Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-02 paper/main.tex
more updates on paper

\newcommand{\true}{\textsc{true}}
\newcommand{\false}{\textsc{false}}
\newcommand{\NULL}{\textsc{null}}
\newcommand{\NA}{---}
\newcommand{\assign}{\ensuremath{:=}}
\newcommand{\procedure}[1]{\textsf{#1}}
\newcommand{\Inf}{\ensuremath{\infty}}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

Kendall's-$\tau$ distance is the metric for rankings that counts the pairs of items ranked in different order in both permutations:
%
\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}

An equivalent definition for the Kendall's-$\tau$ distance counts the number of adjacent swaps that have to be made to convert $\sigma_1^{-1}$ into $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}. \ekhine{ojo, aqui lo dice bien o  mal?}{}

\subsection{Distributions over permutations}
Distributions over permutations are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

In this section we describe our main contribution: Unbalanced Mallows model
(UMM), a novel algorithm for the optimization of a black-box function over
permutation spaces. This algorithm is a probabilistic method based on the
Mallows model and uBorda learning.

\ekhine{10. In Algorithm 1 'S' and 'F' are *sets* (thus, unordered). Therefore, when adding sigma and f(sigma) the correspondence between the two elements seems to be lost. This mapping would be better represented as a set of *pairs* < sigma, f(sigma) >. You always add these elements to both sets at the same time in step 8, anyway.
}{}

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}\Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
    \State $\Sigma \assign$ sample $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}\Comment{Generate $\beta$ permutations}
    \State $\sigma\assign  \argmax_{\pi'\in \Sigma}\min_{ \pi \in S} d(\pi', \pi)   $ \Comment{Select the farthest from $S$}\label{line:distance}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

>>>>

\newcommand{\true}{\textsc{true}}
\newcommand{\false}{\textsc{false}}
\newcommand{\NULL}{\textsc{null}}
\newcommand{\NA}{---}
\newcommand{\assign}{\ensuremath{:=}}
\newcommand{\procedure}[1]{\textsf{#1}}
\newcommand{\Inf}{\ensuremath{\infty}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

Kendall's-$\tau$ distance is the metric for rankings that counts the pairs of items ranked in different order in two permutations:
%
\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}

An equivalent definition for the Kendall's-$\tau$ distance counts the number of adjacent swaps that have to be made to convert $\sigma_1^{-1}$ into $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}. \EKHINE{ojo, aqui lo dice bien o  mal?}\MANUEL{Habría que mirarlo bien!}

\subsection{Distributions over permutations}
Distributions over permutations are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance (Eq.~\ref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

In this section we describe our main contribution: Unbalanced Mallows model
(UMM), a novel algorithm for the optimization of an expensive black-box function over
permutation spaces. This algorithm is a probabilistic method based on the
Mallows model and uBorda learning.

\EKHINE{10. In Algorithm 1 'S' and 'F' are *sets* (thus, unordered). Therefore, when adding sigma and f(sigma) the correspondence between the two elements seems to be lost. This mapping would be better represented as a set of *pairs* < sigma, f(sigma) >. You always add these elements to both sets at the same time in step 8, anyway.
}

\newcommand{\tuple}[1]{\ensuremath{\langle #1\rangle}}

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model Algorithm}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget,\newline\hspace*{5ex} \manuel{$\rho$: learning rate,}{$r_1, r_2$: parameters controlling the learning rate} 
    % \\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign \{\}$
    \For{$i \assign 0$ \textbf{to} $\minit$}
    \State $\sigma \assign \text{generate uniformly at random}$
    \State $S \assign S \cup \tuple{\sigma, f(\sigma)}$\Comment{Evaluate it}
    \EndFor
    %\State $S \assign $ generate  $\minit$ permutations uniformly at random
    %\State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}
    \Statex \Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
    \State $\Sigma \assign$ sample $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    % \Comment{Generate $\beta$ permutations}
    \State $\sigma\assign  \argmax_{\pi'\in \Sigma}\min_{ \pi \in S} d(\pi', \pi)   $ \label{line:distance}
    \Statex \Comment{Select the farthest from $S$}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \tuple{\sigma, f(\sigma)}$\Comment{Evaluate it}
    \EndFor 
    \State \Return $\tuple{\sigma^\ast, f(\sigma^\ast)}\assign \argmin_{\tuple{\sigma, f(\sigma)}\in S} f(\sigma)$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}
